Random number generation and applications on GPU’s in R?

### 2021-09-28 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
add likfit for Betas matrix given
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2876 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\subsection{Creating streams}
In \proglang{clRNG} library, a stream object is a structure that contains three states (see the first column in the \proglang{R} output below): the current state of the stream, the initial state (or seed) of the stream , and the initial state of the current substream. The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host computer using the MRG31k3p RNG. If we want to use the streams in work items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work item picks streams from the global memory and copies their current states only to its private memory. In \pkg{gpuRandom}, to avoid unnecessary data transfer between host and device, we are able to create streams both on the host computer and on the GPU device. \fct{gpuRandom::CreateStreamsCpu} is an \proglang{R} interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on the host as \proglang{R} matrices, so that users can easily view the streams. \fct{gpuRandom::CreateStreamsGpu} is adapted from \fct{clrngMrg31k3pCreateStreams} with the advantage of creating streams directly on GPU device. From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{gpuRandom::CreateStreamsGpu} is more efficient and more frequently used. In the following, we demonstrate the usage of these two functions by generating 4 streams on the host and on the device respectively.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# Create streams on host
streams_on_CPU<-gpuRandom::CreateStreamsCpu(Nstreams=4)
t(streams_on_CPU)
##               [,1]       [,2]       [,3]       [,4]
## current.g1.1 12345  336690377  502033783  739421137
## current.g1.2 12345  597094797 1322587635 1475938232
## current.g1.3 12345 1245771585 1964121530  730262207
## current.g2.1 12345   85196284 1949818481 1630192198
## current.g2.2 12345  523477687 1607232546  324551134
## current.g2.3 12345 2094976052 1462898381  795289868
## initial.g1.1 12345  336690377  502033783  739421137
## initial.g1.2 12345  597094797 1322587635 1475938232
## initial.g1.3 12345 1245771585 1964121530  730262207
## initial.g2.1 12345   85196284 1949818481 1630192198
## initial.g2.2 12345  523477687 1607232546  324551134
## initial.g2.3 12345 2094976052 1462898381  795289868
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{gpuRandom::CreateStreamsCpu} has only one argument \code{Nstreams}, which is the number of streams to create. For the MRG31k3p RNG, each state of the stream object is comprised of six 31-bit integers, the state of the first stream (seed of the \proglang{clRNG} library) is by default ``12345''. By setting a seed same as the MRG31k3p default seed in \fct{gpuRandom::CreateStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# Create streams on GPU
seed <- c(12345, 12345, 12345, 12345, 12345, 12345)
streams1 <- gpuRandom::CreateStreamsGpu(seedR=seed, Nstreams=4, keepInitial=TRUE)
streams_from_GPU <- as.matrix(streams1)
streams2 = deepcopy(streams1)
\end{verbatim}
\end{kframe}
\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
t(streams_from_GPU)   
##               [,1]       [,2]       [,3]       [,4]
## current.g1.1 12345  336690377  502033783  739421137
## current.g1.2 12345  597094797 1322587635 1475938232
## current.g1.3 12345 1245771585 1964121530  730262207
## current.g2.1 12345   85196284 1949818481 1630192198
## current.g2.2 12345  523477687 1607232546  324551134
## current.g2.3 12345 2094976052 1462898381  795289868
## initial.g1.1 12345  336690377  502033783  739421137
## initial.g1.2 12345  597094797 1322587635 1475938232
## initial.g1.3 12345 1245771585 1964121530  730262207
## initial.g2.1 12345   85196284 1949818481 1630192198
## initial.g2.2 12345  523477687 1607232546  324551134
## initial.g2.3 12345 2094976052 1462898381  795289868
\end{verbatim}
\end{kframe}
\end{knitrout}
The \fct{gpuRandom::CreateStreamsGpu} has three arguments:
\begin{itemize}
  \item \code{seedR}: an R vector of length 6, which is the seed of streams. Default is set to a vector of six ``12345''.
  \item \code{Nstreams}: number of streams to create.
  \item \code{keepInitial}: logical specifying whether to keep the seed in generated streams. Default is set to TRUE.
\end{itemize}
\fct{CreateStreamsGpu} in \pkg{gpuRandom} are similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \fct{set.seed} function in \proglang{R}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \pkg{gpuRandom}, with the \code{seedR} argument, we are able to produce exactly the same sequence of streams as many times as we want.

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the runtime between using \fct{stats::rnorm} and \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $512 \times 128$ work-items is around 120 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(gpuRandom::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.056   0.008   0.064
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.932   0.942   7.870
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1e6)$p.value
## [1] 0.4038036
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1e6))
##    user  system elapsed 
##  14.591   0.006  14.590
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for one million simulations, has a p-value of 0.4035.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
almost.1 <- 1 + 64 * .Machine$double.eps
observed_m= -gpuRandom::logfactSum(month, c(16,16))/almost.1

month_GPU<-vclMatrix(month,type="integer")
result_monthgpu <- gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(128,128))
result_monthgpu$simNum
## [1] 1000064
result_monthgpu$counts
## [1] 403788
result_monthgpu$p.value
## [1] 0.4037628
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 999424, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.40375, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40380.  
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(128,128)))
##    user  system elapsed 
##   0.793   0.000   0.792
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{gpuRandom::fisher.sim} takes about 0.32 seconds, \pkg{gpuRandom} accounts for 2.2\% of the elapsed time on CPU.

@@

\end{knitrout}
We perform 10 million simulations for the ``week'' table.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=1e7)$p.value
## [1] 0.0001208
system.time(fisher.test(week,simulate.p.value = TRUE,B=1e7))
##    user  system elapsed 
##  90.691   0.126  90.750
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001208, which should require a larger number of simulations to get a more accurate p-value.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
observed_w= -gpuRandom::logfactSum(week, c(16,16))/almost.1

week_GPU<-gpuR::vclMatrix(week,type="integer")
result_weekgpu<-gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = Nglobal)
result_weekgpu$simNum
## NULL
result_weekgpu$counts
## [1] 1224
result_weekgpu$p.value
## [1] 0.0001225
\end{verbatim}
\end{kframe}
\end{knitrout}
With about ten million simulations, we get 1223 cases and a p-value around 0.0001224705. We summarized the results in Table \ref{tab:summary}.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = Nglobal))
##    user  system elapsed 
##   1.899   0.000   1.898
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes 90.1 seconds, while \fct{fisher.sim} takes about 1.90 seconds, the elapsed time is decreased to about 20\% of the time taken on CPU.

>>>>

\subsection{Creating streams}
In \proglang{clRNG} library, a stream object is a structure that contains three states (see the first column in the \proglang{R} output below): the current state of the stream, the initial state (or seed) of the stream , and the initial state of the current substream. The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host computer using the MRG31k3p RNG. If we want to use the streams in work items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work item picks streams from the global memory and copies their current states only to its private memory. In \pkg{gpuRandom}, to avoid unnecessary data transfer between host and device, we are able to create streams both on the host computer and on the GPU device. \fct{gpuRandom::CreateStreamsCpu} is an \proglang{R} interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on the host as \proglang{R} matrices, so that users can easily view the streams. \fct{gpuRandom::CreateStreamsGpu} is adapted from \fct{clrngMrg31k3pCreateStreams} with the advantage of creating streams directly on GPU device. From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{gpuRandom::CreateStreamsGpu} is more efficient and more frequently used. In the following, we demonstrate the usage of these two functions by generating 4 streams on the host and on the device respectively.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# Create streams on host
streams_on_CPU<-gpuRandom::CreateStreamsCpu(Nstreams=4)
t(streams_on_CPU)
##                 [,1]       [,2]       [,3]       [,4]
## current.g1.1   12345  336690377  502033783  739421137
## current.g1.2   12345  597094797 1322587635 1475938232
## current.g1.3   12345 1245771585 1964121530  730262207
## current.g2.1   12345   85196284 1949818481 1630192198
## current.g2.2   12345  523477687 1607232546  324551134
## current.g2.3   12345 2094976052 1462898381  795289868
## initial.g1.1   12345  336690377  502033783  739421137
## initial.g1.2   12345  597094797 1322587635 1475938232
## initial.g1.3   12345 1245771585 1964121530  730262207
## initial.g2.1   12345   85196284 1949818481 1630192198
## initial.g2.2   12345  523477687 1607232546  324551134
## initial.g2.3   12345 2094976052 1462898381  795289868
## substream.g1.1 12345  336690377  502033783  739421137
## substream.g1.2 12345  597094797 1322587635 1475938232
## substream.g1.3 12345 1245771585 1964121530  730262207
## substream.g2.1 12345   85196284 1949818481 1630192198
## substream.g2.2 12345  523477687 1607232546  324551134
## substream.g2.3 12345 2094976052 1462898381  795289868
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{gpuRandom::CreateStreamsCpu} has only one argument \code{Nstreams}, which is the number of streams to create. For the MRG31k3p RNG, each state of the stream object is comprised of six 31-bit integers, the state of the first stream (seed of the \proglang{clRNG} library) is by default ``12345''. By setting a seed same as the MRG31k3p default seed in \fct{gpuRandom::CreateStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# Create streams on GPU
seed <- c(12345, 12345, 12345, 12345, 12345, 12345)
streams1 <- gpuRandom::CreateStreamsGpu(seedR=seed, Nstreams=4, keepInitial=TRUE)
streams_from_GPU <- as.matrix(streams1)
streams2 = deepcopy(streams1)
\end{verbatim}
\end{kframe}
\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
t(streams_from_GPU)   
##                 [,1]       [,2]       [,3]       [,4]
## current.g1.1   12345  336690377  502033783  739421137
## current.g1.2   12345  597094797 1322587635 1475938232
## current.g1.3   12345 1245771585 1964121530  730262207
## current.g2.1   12345   85196284 1949818481 1630192198
## current.g2.2   12345  523477687 1607232546  324551134
## current.g2.3   12345 2094976052 1462898381  795289868
## initial.g1.1   12345  336690377  502033783  739421137
## initial.g1.2   12345  597094797 1322587635 1475938232
## initial.g1.3   12345 1245771585 1964121530  730262207
## initial.g2.1   12345   85196284 1949818481 1630192198
## initial.g2.2   12345  523477687 1607232546  324551134
## initial.g2.3   12345 2094976052 1462898381  795289868
## substream.g1.1 12345  336690377  502033783  739421137
## substream.g1.2 12345  597094797 1322587635 1475938232
## substream.g1.3 12345 1245771585 1964121530  730262207
## substream.g2.1 12345   85196284 1949818481 1630192198
## substream.g2.2 12345  523477687 1607232546  324551134
## substream.g2.3 12345 2094976052 1462898381  795289868
\end{verbatim}
\end{kframe}
\end{knitrout}
The \fct{gpuRandom::CreateStreamsGpu} has three arguments:
\begin{itemize}
  \item \code{seedR}: an R vector of length 6, which is the seed of streams. Default is set to a vector of six ``12345''.
  \item \code{Nstreams}: number of streams to create.
  \item \code{keepInitial}: logical specifying whether to keep the seed in generated streams. Default is set to TRUE.
\end{itemize}
\fct{CreateStreamsGpu} in \pkg{gpuRandom} are similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \fct{set.seed} function in \proglang{R}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \pkg{gpuRandom}, with the \code{seedR} argument, we are able to produce exactly the same sequence of streams as many times as we want.

@@

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the runtime between using \fct{stats::rnorm} and \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $512 \times 128$ work-items is around 120 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(gpuRandom::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.140   0.004   0.143
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.610   0.862   7.466
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1e6)$p.value
## [1] 0.4042196
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1e6))
##    user  system elapsed 
##  14.578   0.002  14.569
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for one million simulations, has a p-value of 0.4035.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
almost.1 <- 1 + 64 * .Machine$double.eps
observed_m= -gpuRandom::logfactSum(month, c(16,16))/almost.1

month_GPU<-vclMatrix(month,type="integer")
result_monthgpu <- gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(128,128))
result_monthgpu$simNum
## [1] 1015808
result_monthgpu$counts
## [1] 410098
result_monthgpu$p.value
## [1] 0.4037166
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 999424, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.40375, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40380.  
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(128,128)))
##    user  system elapsed 
##   0.702   0.005   0.706
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{gpuRandom::fisher.sim} takes about 0.32 seconds, \pkg{gpuRandom} accounts for 2.2\% of the elapsed time on CPU.

@@

\end{knitrout}
We perform 10 million simulations for the ``week'' table.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=1e7)$p.value
## [1] 0.0001284
system.time(fisher.test(week,simulate.p.value = TRUE,B=1e7))
##    user  system elapsed 
##  90.259   0.150  90.439
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001208, which should require a larger number of simulations to get a more accurate p-value.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
observed_w= -gpuRandom::logfactSum(week, c(16,16))/almost.1

week_GPU<-gpuR::vclMatrix(week,type="integer")
result_weekgpu<-gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = Nglobal)
result_weekgpu$simNum
## [1] 10010624
result_weekgpu$counts
## [1] 1260
result_weekgpu$p.value
## [1] 0.0001259662
\end{verbatim}
\end{kframe}
\end{knitrout}
With about ten million simulations, we get 1223 cases and a p-value around 0.0001224705. We summarized the results in Table \ref{tab:summary}.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = Nglobal))
##    user  system elapsed 
##   4.477   0.009   4.483
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes 90.1 seconds, while \fct{fisher.sim} takes about 1.90 seconds, the elapsed time is decreased to about 20\% of the time taken on CPU.
